---
title: "POLISCI 9590 - Lab0: Introduction to R and RMD"
author: "William Poirier"
date: "2025-09-10"
output: 
  html_document:
      toc: true
      toc_float:
        collapsed: true
      theme: journal
---
<style>
.list-group-item.active, .list-group-item.active:hover, .list-group-item.active:focus {
    z-index: 2;
    color: #ffffff;
    background-color: #4F2683;
    border-color: #4F2683;
}

a {
    color: #4F2683;
    text-decoration: none;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 4. Exercises

1. Install and Load the `carData` package.
```{r}
# install.packages("carData")
library(carData)

```

2. How many columns and rows does the `TitanicSurvival` dataset from `carData` have?
```{r}
dim(TitanicSurvival)
nrow(TitanicSurvival)
ncol(TitanicSurvival)
```

3. What is the mean age of known passenger of the titanic? (Hint: use the `mean()` function and one of its arguments to ignore the NAs.)
```{r}
mean(TitanicSurvival$age,na.rm=T)
```

4. Create a graph of the distribution of age of the titanic passengers. (Hint: use the `hist()` function.)
```{r}
hist(TitanicSurvival$age)
```
